Entropy stable reduced order modeling of nonlinear conservation laws

### 2019-08-08 docs/ESDG_ROM/main.tex
updating paper

\pdfoutput=1
%\documentclass[review,onefignum,onetabnum,final]{siamart171218}
\documentclass[review]{siamart171218}
%\documentclass{article}

%\usepackage{fullpage}
\usepackage{cite}
%\usepackage{geometry}

\usepackage{amsmath, amsfonts, amssymb}
\DeclareMathOperator*{\argmin}{\arg\,\min}
\DeclareMathOperator*{\argmax}{\arg\,\max}
\usepackage{algorithmic}

\newtheorem*{remark}{Remark}
\theoremstyle{assumption}
\newtheorem{assumption}{Assumption}

@@

\usepackage{hyperref}
\usepackage[titletoc,toc,title]{appendix}
\usepackage{tikz}
%\usepackage{hhline}
\usepackage{graphicx, color, bm}
\usepackage{subfig}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{slopetri} % my own package for drawing triangle slopes

@@

\date{}
\author{Jesse Chan}
\title{Entropy stable reduced order modeling of nonlinear conservation laws}

@@

\maketitle

\begin{abstract}
Reduced order models of nonlinear conservation laws in fluid dynamics do not typically inherit stability properties of the full order model. We discuss the construction of reduced models of nonlinear conservation laws which retain accuracy while inheriting a semi-discrete entropy inequality independently of discretization resolutions and parameters. 
\end{abstract}

\tableofcontents

%\begin{itemize}
%\item Intro
%\item Entropy conservation for FOM
%\item Entropy conservation for POD
%\item Entropy stability via entropy dissipation
%\item Entropy stable hyper-reduction
%\item Proof: accuracy of projected operator
%\item Proof: entropy stability for periodic domain
%\item Proof: conservation if $\bm{1}$ is contained in $\bm{V}_r$
%\item Non-periodic domains and constrained hyper-reduction

%\item Numerical experiments
%\begin{itemize}
%\item Singular values of conservative snapshots vs conservative + entropy variable snapshots.
%\item 1D experiments, shocks, error convergence plots
%\item Effect of different hyper-reduction techniques (optimizing for $V_{\rm test} V_{\rm test}$ and $V_r V_{\rm test}$ vs $V_r V_r$)
%\item Comparison of entropy stable viscosity vs regular hyper-reduced viscosity
%\item Comparison of snapshots from entropy stable FOM vs regular FOM.  
%\item 2D experiments: KH and acoustic pulse
%\end{itemize}
%\end{itemize}

@@

Projection-based model reduction constructs low-dimensional surrogate models for many-query scenarios (e.g., simulations over multiple parameter values) that can be evaluated over a range of parameters at a low \textit{online} cost in exchange for a more expensive \textit{offline} pre-computation step \cite{benner2015survey}.  %Reduced order models (ROMs) can be constructed, for example, by performing a Galerkin projection of the original full order model onto a reduced basis.  
The development of reduced order models (ROMs) is relatively mature for several classes of problems (e.g., linear time-invariant systems, coercive elliptic PDEs).  However, the construction of robust and stable ROMs for transient and convection-dominated problems remains an active area of research \cite{cagniart2019model}.

For certain PDEs, ``structure-preserving'' ROMs provide robustness and stability by preserving energetic properties of the full system at the discrete level.  ROMs which retain either Lagrangian \cite{lall2003structure, carlberg2015preserving} or Hamiltonian structure \cite{polyuga2010structure, benner2012interpolation, gugercin2012structure, peng2016symplectic, chaturantabut2016structure, gong2017structure, afkham2017structure, afkham2018structure} have been constructed by combining Galerkin projection with an appropriate formulation of the full order model, and similar energy-conserving ROMs have been constructed for the incompressible Navier-Stokes equations \cite{farhat2014dimensional, farhat2015structure}.  The construction of structure-preserving ROMs to nonlinear conservation laws in fluid flow, however, remains an open problem.  For example, Galerkin projection yields ROMs which become unstable as the number of reduced basis functions (modes) is increased \cite{bui2007goal, carlberg2013gnat}.  As a result, stabilized discretizations are often employed.  Petrov-Galerkin ROMs, which use an alternative test basis \cite{maday2002blackbox, rozza2007stability, bui2007goal, amsallem2012stabilization, rozza2013reduced, ballarin2015supremizer, carlberg2017galerkin}, are a popular alternative, as are additional residual-based stabilization or dissipation terms \cite{wang2012proper, kalashnikova2014stabilization, caiazzo2014numerical, balajewicz2016minimal}.  Such models improve robustness in practice, though they do not provide a theoretical foundation for stability.

The construction of stable ROMs for the compressible Navier-Stokes equations is complicated by their complex structure.  Thus, practitioners often develop structure-preserving entropy stable ROMs for simplifications of the underlying PDE (such as the time-dependent \textit{linearized} compressible Navier-Stokes equations \cite{barone2009stable, kalashnikova2010stability, serre2012reliable, amsallem2012stabilization, kalashnikova2014construction}) and extrapolate such models to the full nonlinear equations.  A promising alternative is to enforce physical conditions such as such as kinetic energy preservation, which are empirically related to stability \cite{maboudi2018conservative}.  These methods significantly improve the robustness of numerical methods in practice, and can guarantee a discrete entropy inequality for systems of nonlinear conservation laws which yield stable split formulations \cite{fisher2013discretely}.

@@

In this work, we directly constructing entropy stable ROMs for nonlinear conservation laws through a modified Galerkin projection of an appropriate full order model.  The approach taken in this paper combines techniques from entropy stable finite volume schemes \cite{tadmor1987numerical, tadmor2003entropy} and entropy stable summation-by-parts (SBP) discretizations \cite{fisher2013high, carpenter2014entropy, chen2017entropy, crean2018entropy, chan2017discretely, chan2018discretely, chan2019skew} to produce entropy stable reduced order models.  The paper is organized as follows.  Section~\ref{sec:1} discusses the entropy inequality associated with systems of nonlinear conservation laws.  Section~\ref{sec:2} introduces a full order model on 1D periodic domains based on entropy stable finite volume schemes.  Section~\ref{sec:3} describes how to construct an entropy stable reduced basis approximation, while Section~\ref{sec:4} discusses entropy stable hyper-reduction techniques to reduce costs associated with the evaluation of nonlinear terms.  Section~\ref{sec:6} describes how to extend the aforementioned approaches to non-periodic boundary conditions, and Section~\ref{sec:7} describes the extension to higher dimensions.  We conclude in Section~\ref{sec:8} with numerical experiments verifying theoretical results.

\section{Entropy inequalities for nonlinear conservation laws}
\label{sec:1}
Let $\Omega$ denote some domain with boundary $\partial \Omega$.  General nonlinear conservation laws are expressed as a system of partial differential equations (PDEs) 
\begin{equation}
\pd{\bm{u}}{t}  + \sum_{i=1}^d\pd{\bm{f}_i(\bm{u})}{x_i} = 0, \qquad 
S(\bm{u}) \text{ is a convex function}, \qquad
\bm{v}(\bm{u}) = \pd{S}{\bm{u}},
\label{eq:nonlineqs}
\end{equation}
where $\bm{u}\in \mathbb{R}^n$ are the conservative variables, $\bm{f}_j$ are nonlinear fluxes, and $\bm{v}(\bm{u})$ are the \textit{entropy variables}.  By multiplying (\ref{eq:nonlineqs}) by the entropy variables, viscosity solutions \cite{oleinik1957discontinuous, kruvzkov1970first}  of popular fluid systems (e.g., shallow water, compressible Euler and Navier-Stokes \cite{hughes1986new, chen2017entropy}) can be shown to satisfy 
\begin{equation}
\int_{\Omega}\pd{S(\bm{u})}{t}\diff{\bm{x}} + \int_{\partial \Omega} \LRp{\bm{v}^T\bm{f}(\bm{u}) - \psi_i(\bm{u})}n_i \leq 0\label{eq:entropyineq},
\end{equation}
where $n_i$ denotes the $i$th component of the outward normal vector.  
The entropy inequality (\ref{eq:entropyineq}) is the analogue of energy stability for nonlinear conservation laws \cite{mock1980systems, harten1983symmetric}.  
However, due to the use of inexact treatment of nonlinear terms (e.g., from collocation approximations, aliasing errors, or inexact quadrature rules), most numerical methods for nonlinear conservation laws do not satisfy a discrete analogue of this stability condition.

@@

We first rewrite the system (\ref{eq:nonlineqs1d}) in a matrix form which will be more amenable to model reduction.  We define the skew-symmetric matrix $\bm{Q}$ and flux matrix $\bm{F}$ such that
\begin{equation}
\bm{Q} = \frac{1}{2}\begin{bmatrix}
0 & 1 & &\ldots & -1\\
-1 & 0 & 1 &&  \\
& -1 & 0 & 1 &  \\
 & & & \ddots &  \\
1 & &\ldots  & -1 & 0
\end{bmatrix}, \qquad \bm{F}_{ij} = \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j}.
\label{eq:Qmat}
\end{equation}
Note that $\bm{Q}\bm{1} = \bm{0}$, the matrix $\bm{F}$ is symmetric due to flux symmetry, and the diagonal of $\bm{F}$ is equal to $\bm{f}(\bm{u})$ due to flux consistency.  For a scalar nonlinear conservation law, the matrix-based formulation of (\ref{eq:nonlineqs1d}) is then equivalent to
\begin{align}
\Delta x \td{\bm{u}_h}{t} + 2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1} = \bm{0}.
\label{eq:es}
\end{align}
where $\circ$ denotes the Hadamard product, $\bm{1}$ denotes the vector of all ones, and each entry of $\bm{u}_h$ corresponds to a point $\bm{x}_i$ in the domain.  In this setting, $\bm{Q}\circ\bm{F}$ extracts and takes linear combinations of nonlinear flux interactions between different nodal values of $(\bm{u}_h)_i$ and $(\bm{u}_h)_j$.  The discretization of the nonlinear flux term using $\LRp{\bm{Q}\circ \bm{F}}\bm{1}$ is commonly referred to as \textit{flux differencing} \cite{carpenter2014entropy, gassner2016split, chen2017entropy, crean2018entropy, chan2017discretely}.\footnote{The factor of 1/2 present in the definition of $\bm{Q}$ is to make (\ref{eq:es}) consistent with the entropy stable SBP literature.  The factor of 2 in (\ref{eq:es}) can also be derived from the chain rule \cite{chen2017entropy, crean2018entropy}.  }

@@

We next show that the full order model satisfies a discrete conservation of entropy.  
\begin{theorem}
\label{thm:ecfom}
Let $\bm{u}_h(t)$ be a solution of (\ref{eq:essys}) for which the entropy $S(\bm{u}_h)$ is convex at each time $t$.  Then, $\bm{u}_h$ satisfies the semi-discrete conservation of entropy 
\[
\Delta x \bm{1}^T\td{S(\bm{u}_h)}{t} = 0.
\]
\end{theorem}
\begin{proof}
The proof can be found in the literature \cite{tadmor1987numerical, tadmor2003entropy, tadmor2016entropy}.  We present an alternative proof which utilizes matrix properties of entropy stable summation by parts (SBP) schemes \cite{carpenter2014entropy, gassner2016split, chen2017entropy, crean2018entropy, chan2017discretely}.  We test (\ref{eq:es}) with the vector of entropy variables $\bm{v}_h = \bm{v}(\bm{u}_h)$
\[
\Delta x \bm{v}_h^T\td{\bm{u}_h}{t} + \bm{v}_h^T2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1} = 0.
\]
Assuming the chain rule in time, 
\begin{align*}
\bm{v}_h^T\td{\bm{u}_h}{t} &= {\td{S(\bm{u}_h)}{\bm{u}}}^T\td{\bm{u}_h}{t} = \sum_j \LRp{\td{S(\bm{u}_h)}{\bm{u}}}_j^T\td{\LRp{\bm{u}_h}_j}{t} \\
&=  \sum_j \td{S(\LRp{\bm{u}_h}_j)}{t}  =  \bm{1}^T\td{S(\bm{u}_h)}{t}.
\end{align*}
Using skew-symmetry of $\bm{Q}$, the flux term yields
\begin{align*}
\bm{v}_h^T2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1} &= \sum_{ij} \LRp{\bm{v}_h}_i^T \bm{Q}_{ij} \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j} \\
&= \sum_{ij} \LRp{\bm{Q}_{ij}-\bm{Q}_{ji}} \LRp{\bm{v}_h}_i^T  \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j}.
\end{align*}
Rearranging indices and using the symmetry of $\bm{f}_S(\bm{u}_L,\bm{u}_R) = \bm{f}_S(\bm{u}_R,\bm{u}_L)$ exposes the entropy conservation condition in the sum
\begin{align*}
&\sum_{ij} \LRp{\bm{Q}_{ij}-\bm{Q}_{ji}} \LRp{\bm{v}_h}_i^T \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j} \\
=&\sum_{ij} \bm{Q}_{ij} \LRp{\LRp{\bm{v}_h}_i-\LRp{\bm{v}_h}_j}^T  \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j}\\
=&\sum_{ij} \bm{Q}_{ij} \LRp{\psi\LRp{\LRp{\bm{u}_h}_i} - \psi(\LRp{\bm{u}_h}_j)} = \bm{\psi}^T\bm{Q}\bm{1}-\bm{1}^T\bm{Q}\bm{\psi} = 0,
\end{align*}
where we have used that $\bm{Q}=-\bm{Q}^T$ and $\bm{Q}\bm{1} = \bm{0}$ in the final step.

@@

\subsection{Viscosity and entropy dissipation}

Theorem~\ref{thm:ecfom} shows that the formulation (\ref{eq:essys}) preserves a semi-discrete conservation of entropy.  However, in the presence of shock discontinuities, entropy should be dissipated instead of conserved.  To mimic this at the semi-discrete level, we add entropy dissipation to (\ref{eq:es}) through appropriate viscosity terms.  For example, it was shown in \cite{tadmor2006entropy} that a centered approximation of the Navier-Stokes viscosity is entropy dissipative.  In this work, we apply a simple Laplacian artificial viscosity to each component of the solution \cite{upperman2019entropy}
\begin{equation}
\Delta x \td{\bm{u}_h}{t} + 2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1} + \epsilon \bm{K}\bm{u}_h= 0, \qquad \bm{K} = \frac{1}{\Delta x^2} \begin{bmatrix}
1 & -1 & & \\
-1 & 2 & \ddots &\\
 & \ddots & \ddots & -1\\
 &  & -1& 1\\
\end{bmatrix}
\label{eq:esvisc}
\end{equation}
where $\epsilon$ is the visosity coefficient.  This choice of dissipation simplifies descriptions of how to treat viscous regularizations of the solution in an entropy stable fashion.  Future work will explore both physically relevant viscosities and more nuanced artificial dissipation mechanisms.

@@

\bm{V}^T\bm{V}\td{\bm{u}_N}{t} + 2\bm{V}^T\LRp{\bm{Q}\circ \bm{F}}\bm{1} = 0.

@@

\bm{V}^T\bm{V}\td{\bm{u}_N}{t} + 2\bm{V}^T\LRp{\bm{Q}\circ \bm{F}}\bm{1} = 0, \label{eq:esrom}\\

@@

@@

The result follows from testing with $\bm{v}_N$ and applying (\ref{eq:esromtime}).  The remaining steps are identical to those of Theorem~\ref{thm:ecfom}.  

@@

\subsubsection{Choices of test basis}

We have not yet specified the choice of test basis $\bm{V}_t$.  Common sense suggests that the span of the test basis should include the reduced basis $\bm{V}$.  Theorem~\ref{thm:Qt} implies that the span of the test basis should also contain $\bm{1}$.  However, these two conditions are insufficient to ensure accuracy for solutions of nonlinear conservation laws.  Consider, for example, the 1D periodic Burgers' equation with initial condition 
\[
u(x,0) = -\sin(\pi x).
\]
The solution $u(x,t)$ is a decaying stationary shock which is anti-symmetric across the origin at all times $t$.  The POD modes are thus also anti-symmetric across the origin; however, their derivatives $\bm{Q}\bm{V}$ are symmetric across the origin, as shown in Figure~\ref{fig:modeQ}.  As a result, since the entries of $\bm{V}^T\bm{Q}\bm{V}$ are inner products of symmetric and anti-symmetric functions, $\bm{V}^T\bm{Q}\bm{V} \approx \bm{0}$.
\begin{figure}[!h]
\centering
\subfloat[Solution snapshots]{\includegraphics[width=.3\textwidth]{figs/burg_snaps.png}}
\hspace{.5em}
\subfloat[Modes (columns of $\bm{V}$)]{\includegraphics[width=.3\textwidth]{figs/burg3modes.png}}
\hspace{.5em}
\subfloat[Mode derivatives $\bm{Q}\bm{V}$]{\includegraphics[width=.3\textwidth]{figs/burg3modes_diff.png}}
\caption{Solution snapshots, three POD modes, and mode derivatives for a shock solution of the viscous periodic Burgers' equation. }
\label{fig:modeQ}
\end{figure}

@@

\subsection{Entropy dissipation}

>>>>

\pdfoutput=1
%\documentclass[review,onefignum,onetabnum,final]{siamart171218}
%\documentclass[review]{siamart171218}
\documentclass[preprint,10pt]{elsarticle}

\usepackage{fullpage}
%\usepackage{cite}
%\usepackage{geometry}[scale=.8]

\usepackage[colorlinks=true]{hyperref}

\usepackage{amsmath,amssymb,amsfonts,amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{lemma}
\newtheorem{lemma}{Lemma}
\newtheorem*{remark}{Remark}
\theoremstyle{theorem}
\newtheorem{theorem}{Theorem}
\theoremstyle{assumption}
\newtheorem{assumption}{Assumption}

\DeclareMathOperator*{\argmin}{\arg\,\min}
\DeclareMathOperator*{\argmax}{\arg\,\max}
\usepackage{algorithm}
\usepackage{algorithmic}

%\newtheorem*{remark}{Remark}
%\theoremstyle{assumption}
%\newtheorem{assumption}{Assumption}

@@

\usepackage[titletoc,toc,title]{appendix}
\usepackage{tikz}
%\usepackage{hhline}
\usepackage{graphicx, color, bm}
\usepackage{subfig}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
%\usepackage{slopetri} % my own package for drawing triangle slopes

@@

%\date{}
%\author{Jesse Chan}

@@

\begin{frontmatter}
\title{Entropy stable reduced order modeling of nonlinear conservation laws}

\author[rice]{Jesse Chan\corref{cor1}}
\ead{Jesse.Chan@caam.rice.edu}
\address[rice]{Department of Computational and Applied Mathematics, Rice University, 6100 Main St, Houston, TX, 77005}

\begin{abstract}
Reduced order models of nonlinear conservation laws in fluid dynamics do not typically inherit stability properties of the full order model. We construct projection-based reduced models of nonlinear conservation laws based on flux differencing which retain accuracy while inheriting a semi-discrete entropy inequality independently of discretization resolutions and parameters. 
\end{abstract}
\end{frontmatter}

%\maketitle

%\tableofcontents

@@

Projection-based model reduction constructs low-dimensional surrogate models for many-query scenarios (e.g., simulations over multiple parameter values) that can be evaluated over a range of parameters at a low \textit{online} cost in exchange for a more expensive \textit{offline} pre-computation step \cite{benner2015survey}.  The development of reduced order models (ROMs) is relatively mature for several classes of problems (e.g., linear time-invariant systems, coercive elliptic PDEs).  However, the construction of robust and stable ROMs for transient and convection-dominated problems remains an active area of research \cite{cagniart2019model}.

For certain PDEs, ``structure-preserving'' ROMs provide robustness and stability by preserving energetic properties of the full system at the discrete level.  ROMs which retain either Lagrangian \cite{lall2003structure, carlberg2015preserving} or Hamiltonian structure \cite{benner2012interpolation, gugercin2012structure, peng2016symplectic, chaturantabut2016structure, gong2017structure, afkham2017structure, afkham2018structure} have been constructed by combining Galerkin projection with an appropriate formulation of the full order model, and similar energy-conserving ROMs have been constructed for the incompressible Navier-Stokes equations \cite{farhat2014dimensional, farhat2015structure}.  The construction of structure-preserving ROMs to nonlinear conservation laws in fluid flow, however, remains an open problem.  For example, Galerkin projection yields ROMs which become unstable as the number of reduced basis functions (modes) is increased \cite{bui2007goal, carlberg2013gnat}.  As a result, stabilized discretizations are often employed.  Petrov-Galerkin ROMs, which use an alternative test basis \cite{maday2002blackbox, rozza2007stability, bui2007goal, serre2012reliable, amsallem2012stabilization, rozza2013reduced, ballarin2015supremizer, carlberg2017galerkin}, are a popular alternative, as are additional residual-based stabilization or dissipation terms \cite{wang2012proper, kalashnikova2014stabilization, caiazzo2014numerical, balajewicz2016minimal}.  Such models improve robustness in practice, though they do not provide a theoretical foundation for stability.

The construction of stable ROMs for the compressible Navier-Stokes equations is complicated by their complex structure.  Thus, practitioners often develop structure-preserving entropy stable ROMs for simplifications of the underlying PDE (such as the time-dependent \textit{linearized} compressible Navier-Stokes equations \cite{barone2009stable, kalashnikova2010stability, amsallem2012stabilization, kalashnikova2014construction}) and extrapolate such models to the full nonlinear equations.  A promising alternative is to enforce physical conditions such as such as kinetic energy preservation, which are empirically related to stability \cite{maboudi2018conservative}.  These methods significantly improve the robustness of numerical methods in practice, and can guarantee a discrete entropy inequality for systems of nonlinear conservation laws which yield stable split formulations \cite{fisher2013discretely}.

@@

In this work, we directly constructing entropy stable ROMs for nonlinear conservation laws through a modified Galerkin projection of an appropriate full order model.  The approach taken in this paper combines techniques from entropy stable finite volume schemes \cite{tadmor1987numerical, tadmor2003entropy} and entropy stable summation-by-parts (SBP) discretizations \cite{fisher2013high, carpenter2014entropy, chen2017entropy, crean2018entropy, chan2017discretely, chan2019skew} to produce entropy stable reduced order models.

The paper is organized as follows.  Section~\ref{sec:1} discusses the entropy inequality associated with systems of nonlinear conservation laws.  Section~\ref{sec:2} introduces a full order model on 1D periodic domains based on entropy stable finite volume schemes.  Section~\ref{sec:3} describes how to construct an entropy stable reduced basis approximation, while Section~\ref{sec:4} discusses entropy stable hyper-reduction techniques to reduce costs associated with the evaluation of nonlinear terms.  Section~\ref{sec:6} describes how to extend the aforementioned approaches to non-periodic boundary conditions, and Section~\ref{sec:7} describes the extension to higher dimensions.  We conclude in Section~\ref{sec:8} with numerical experiments verifying theoretical results.

\section{Entropy inequalities for nonlinear conservation laws}
\label{sec:1}
Let $\Omega$ denote some domain with boundary $\partial \Omega$.  General nonlinear conservation laws are expressed as a system of partial differential equations (PDEs) 
\begin{equation}
\pd{\bm{u}}{t}  + \sum_{i=1}^d\pd{\bm{f}_i(\bm{u})}{x_i} = 0, \qquad 
S(\bm{u}) \text{ is a convex function}, \qquad
\bm{v}(\bm{u}) = \pd{S}{\bm{u}},
\label{eq:nonlineqs}
\end{equation}
where $\bm{u}\in \mathbb{R}^n$ are the conservative variables, $\bm{f}_j$ are nonlinear fluxes, and $\bm{v}(\bm{u})$ are the \textit{entropy variables}.  By multiplying (\ref{eq:nonlineqs}) by the entropy variables, viscosity solutions \cite{oleinik1957discontinuous, kruvzkov1970first}  of popular fluid systems (e.g., shallow water, compressible Euler and Navier-Stokes \cite{hughes1986new, chen2017entropy}) can be shown to satisfy 
\begin{equation}
\int_{\Omega}\pd{S(\bm{u})}{t}\diff{\bm{x}} + \sum_{i=1}^d \int_{\partial \Omega} \LRp{\bm{v}^T\bm{f}_i(\bm{u}) - \psi_i(\bm{u})}n_i \leq 0\label{eq:entropyineq},
\end{equation}
where $n_i$ denotes the $i$th component of the outward normal vector.  
The entropy inequality (\ref{eq:entropyineq}) is the analogue of energy stability for nonlinear conservation laws \cite{mock1980systems, harten1983symmetric}.  
However, due to the use of inexact treatment of nonlinear terms (e.g., from collocation approximations or inexact quadrature), most numerical methods for nonlinear conservation laws do not satisfy a discrete analogue of this stability condition.

@@

We first rewrite the system (\ref{eq:nonlineqs1d}) in a matrix form which will be more amenable to model reduction.  We define the skew-symmetric differentiation matrix $\bm{Q}$ and flux matrix $\bm{F}$ such that
\begin{equation}
\bm{Q} = \frac{1}{2}\begin{bmatrix}
0 & 1 & &\ldots & -1\\
-1 & 0 & 1 &&  \\
& -1 & 0 & 1 &  \\
 & & & \ddots &  \\
1 & &\ldots  & -1 & 0
\end{bmatrix}, \qquad \bm{F}_{ij} = \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j}.
\label{eq:Qmat}
\end{equation}
Note that $\bm{Q}\bm{1} = \bm{0}$, that the matrix $\bm{F}$ is symmetric (due to symmetry of $\bm{f}_S$), and that the diagonal of $\bm{F}$ is equal to $\bm{f}(\bm{u})$ due to flux consistency.  For a scalar nonlinear conservation law, the matrix-based formulation of (\ref{eq:nonlineqs1d}) is then equivalent to
\begin{align}
\Delta x \td{\bm{u}_h}{t} + 2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1} = \bm{0}.
\label{eq:es}
\end{align}
where $\circ$ denotes the Hadamard product, $\bm{1}$ denotes the vector of all ones, and each entry of $\bm{u}_h$ corresponds to a point $\bm{x}_i$ in the domain.  In this setting, $\bm{Q}\circ\bm{F}$ extracts and takes linear combinations of nonlinear flux interactions between different nodal values of $(\bm{u}_h)_i$ and $(\bm{u}_h)_j$.  The discretization of the nonlinear flux term using $\LRp{\bm{Q}\circ \bm{F}}\bm{1}$ is commonly referred to as \textit{flux differencing} \cite{carpenter2014entropy, gassner2016split, chen2017entropy, crean2018entropy, chan2017discretely}.\footnote{The factor of 1/2 present in the definition of $\bm{Q}$ is to make (\ref{eq:es}) consistent with the entropy stable SBP literature.  The factor of 2 in (\ref{eq:es}) can also be derived from the chain rule \cite{chen2017entropy, crean2018entropy}.  }

@@

We next show that the full order model satisfies a discrete conservation of entropy.  
\begin{theorem}
\label{thm:ecfom}
Let $\bm{u}_h(t)$ be a solution of (\ref{eq:essys}) for which the entropy $S(\bm{u}_h)$ is convex at each time $t$.  Then, $\bm{u}_h$ satisfies the semi-discrete conservation of entropy 
\[
\Delta x \bm{1}^T\td{S(\bm{u}_h)}{t} = 0.
\]
\end{theorem}
\begin{proof}
The proof can be found in the literature \cite{tadmor1987numerical, tadmor2003entropy, tadmor2016entropy}.  We present an alternative proof which utilizes matrix properties of entropy stable summation by parts (SBP) schemes \cite{carpenter2014entropy, gassner2016split, chen2017entropy, crean2018entropy, chan2017discretely}.  We test (\ref{eq:es}) with the vector of entropy variables $\bm{v}_h = \bm{v}(\bm{u}_h)$
\[
\Delta x \bm{v}_h^T\td{\bm{u}_h}{t} + \bm{v}_h^T2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1} = 0.
\]
Assuming continuity in time and using the definition of the entropy variables in (\ref{eq:nonlineqs}), $\bm{v}_h^T\td{\bm{u}_h}{t}$ simplifies to
\begin{align*}
{\td{S(\bm{u}_h)}{\bm{u}}}^T\td{\bm{u}_h}{t} = \sum_j \LRp{\td{S(\bm{u}_h)}{\bm{u}}}_j^T\td{\LRp{\bm{u}_h}_j}{t} =  \sum_j \td{S(\LRp{\bm{u}_h}_j)}{t}  =  \bm{1}^T\td{S(\bm{u}_h)}{t}.
\end{align*}
Using skew-symmetry of $\bm{Q}$, the flux term $\bm{v}_h^T2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1}$ yields
\begin{align*}
\sum_{ij} \LRp{\bm{v}_h}_i^T \bm{Q}_{ij} \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j} = \sum_{ij} \LRp{\bm{Q}_{ij}-\bm{Q}_{ji}} \LRp{\bm{v}_h}_i^T  \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j}.
\end{align*}
Rearranging indices and using the symmetry of $\bm{f}_S(\bm{u}_L,\bm{u}_R) = \bm{f}_S(\bm{u}_R,\bm{u}_L)$ exposes the entropy conservation condition in the sum
\begin{align*}
&\sum_{ij} \LRp{\bm{Q}_{ij}-\bm{Q}_{ji}} \LRp{\bm{v}_h}_i^T \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j} \\
=&\sum_{ij} \bm{Q}_{ij} \LRp{\LRp{\bm{v}_h}_i-\LRp{\bm{v}_h}_j}^T  \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j}\\
=&\sum_{ij} \bm{Q}_{ij} \LRp{\psi\LRp{\LRp{\bm{u}_h}_i} - \psi(\LRp{\bm{u}_h}_j)} = \bm{\psi}^T\bm{Q}\bm{1}-\bm{1}^T\bm{Q}\bm{\psi} = 0,
\end{align*}
where we have used that $\bm{Q}=-\bm{Q}^T$ and $\bm{Q}\bm{1} = \bm{0}$ in the final step.

@@

\subsection{Viscosity and entropy dissipation} 
\label{sec:entropydissipation}
Theorem~\ref{thm:ecfom} shows that the formulation (\ref{eq:essys}) preserves a semi-discrete conservation of entropy.  However, in the presence of shock discontinuities, entropy should be dissipated instead of conserved.  To mimic this at the semi-discrete level, we add entropy dissipation to (\ref{eq:es}) through appropriate viscosity terms.  For example, it was shown in \cite{tadmor2006entropy} that a centered approximation of the Navier-Stokes viscosity is entropy dissipative.  In this work, we apply a simple Laplacian artificial viscosity to each component of the solution \cite{upperman2019entropy}
\begin{equation}
\Delta x \td{\bm{u}_h}{t} + 2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1} + \epsilon \bm{K}\bm{u}_h= 0, \qquad \bm{K} = \frac{1}{\Delta x^2} \begin{bmatrix}
1 & -1 & & \\
-1 & 2 & \ddots &\\
 & \ddots & \ddots & -1\\
 &  & -1& 1\\
\end{bmatrix}
\label{eq:esvisc}
\end{equation}
where $\epsilon$ is the visosity coefficient.  This choice of dissipation simplifies descriptions of how to treat viscous regularizations of the solution in an entropy stable fashion.  Future work will explore both physically relevant viscosities and more nuanced artificial dissipation mechanisms.

@@

\Delta x\bm{V}^T\bm{V}\td{\bm{u}_N}{t} + 2\bm{V}^T\LRp{\bm{Q}\circ \bm{F}}\bm{1} = 0.

@@

\Delta x\bm{V}^T\bm{V}\td{\bm{u}_N}{t} + 2\bm{V}^T\LRp{\bm{Q}\circ \bm{F}}\bm{1} = \bm{0}, \label{eq:esrom}\\

@@

Additionally, if the reduced basis $\bm{V}$ exactly represents $\bm{1}$, solutions of (\ref{eq:esrom}) conserve global averages of the conservative variables.

@@

The conservation of entropy follows from testing with $\bm{v}_N$ and applying (\ref{eq:esromtime}).  The remaining steps are identical to those of Theorem~\ref{thm:ecfom}.  The global conservation results from testing with $\bm{1}$.  Let $\bm{V}\bm{e}=\bm{1}$, where $\bm{e}$ denote coefficients of $\bm{1}$ in the reduced basis matrix.  Then, 
\Delta x\bm{e}^T\bm{V}^T\bm{V}\td{\bm{u}_N}{t} + 2\bm{e}^T\bm{V}^T\LRp{\bm{Q}\circ \bm{F}}\bm{1}  = \Delta x\bm{1}^T\td{\LRp{\bm{V}\bm{u}_N}}{t} + 2\bm{1}^T\LRp{\bm{Q}\circ \bm{F}}\bm{1} = 0.
Since $\bm{Q}$ is skew-symmetric and $\bm{F}$ is symmetric, $\bm{Q} \circ \bm{F}$ is skew-symmetric and $\bm{1}^T\LRp{\bm{Q}\circ \bm{F}}\bm{1} = \bm{0}$, 
\Delta x\bm{1}^T\td{\LRp{\bm{V}\bm{u}_N}}{t} = 0,
which is a discrete statement of conservation.

@@

\subsubsection{Choice of test basis}

We have not yet specified the choice of test basis $\bm{V}_t$.  The span of the test basis should include the reduced basis $\bm{V}$, and Lemma~\ref{thm:Qt} implies that the span of the test basis should also contain $\bm{1}$.  However, these two conditions are insufficient to ensure accuracy for solutions of nonlinear conservation laws.  Consider, for example, the 1D periodic Burgers' equation with initial condition 
\[
u(x,0) = -\sin(\pi x).
\]
The solution $u(x,t)$ is a decaying stationary shock which is anti-symmetric across the origin at all times $t$.  The POD modes are thus also anti-symmetric across the origin; however, their derivatives $\bm{Q}\bm{V}$ are symmetric across the origin, as shown in Figure~\ref{fig:modeQ}.  As a result, since the entries of $\bm{V}^T\bm{Q}\bm{V}$ are inner products of symmetric and anti-symmetric functions, $\bm{V}^T\bm{Q}\bm{V} \approx \bm{0}$.
\begin{figure}[!h]
\centering
\subfloat[Solution snapshots]{\includegraphics[width=.32\textwidth]{figs/burg_snaps.png}}
\hspace{.1em}
\subfloat[Modes (columns of $\bm{V}$)]{\includegraphics[width=.32\textwidth]{figs/burg3modes.png}}
\hspace{.1em}
\subfloat[Mode derivatives $\bm{Q}\bm{V}$]{\includegraphics[width=.32\textwidth]{figs/burg3modes_diff.png}}
\caption{Solution snapshots, three POD modes, and mode derivatives for a shock solution of the viscous periodic Burgers' equation. }
\label{fig:modeQ}
\end{figure}

@@

%\subsection{Entropy dissipation}
%\note{Discuss how to produce entropy dissipative hyper-reduced operators in 2D}.
